### 2012-09-02 sections/design.tex
* introduction   * capturing flows     * elevator pitch blurb     * netflow and ipfix     * wide applicability of flow capture techniques     * flow query language     * NFQL implementation     * paper organization * related work   * tcpdump, wireshark, flow-tools, nfdump, silk   * silk citation, simpleweb citation * flow query lanauage

The pipeline consists of a number of independent processing elements that are
connected to one another using UNIX-based pipes. Each element receives the
content from the previous pipe, performs an operation and pushes it to the
next element in the pipeline. Fig. \ref{fig:nfql-pipeline} shows an
overview of the processing pipeline. A complete description on the semantics
of each element in the pipeline can be found in \cite{vmarinov:2009}

The splitter takes the flow-records data as input in the \texttt{flow-tools}
compatible format. It is responsible to duplicate the input data out to
several branches without any processing whatsoever. This
allows each of the branches to have an identical copy of the flow data to
process it independently.

The filter performs \emph{absolute} filtering on the input flow-records data.
The flow-records that pass the filtering criterion are forwarded to the
grouper, the rest of the flow-records are dropped. The
filter compares separate fields of a flow-record against either a constant
value or a value on a different field of the \emph{same} flow-record. The
filter cannot \emph{relatively} compare two different incoming flow-records

The grouper performs aggregation of the input flow-records data. It consists
of a number of rule modules that correspond to a specific subgroup. A
flow-record in order to be a part of the group should be a part of at-least
one subgroup. A flow-record can be a part of multiple subgroups within a
group. A flow-record cannot be part of multiple groups.
The grouping rules can be either absolute or relative. The newly formed groups
which are passed on to the group filter can also contain meta-information
about the flow-records contained within the group using the aggregate clause
defined as part of the grouper query.

The group-filter performs \emph{absolute} filtering on the input group-records
data. The group-records that pass the filtering criterion are forwarded to the
merger, the rest of the group-records are dropped. The group-filter compares
separate fields (or aggregated fields) of a
flow-record against either a constant value or a value on a different field of
the \emph{same} flow-record. The group-filter cannot \emph{relatively} compare
two different incoming group-records

The merger performs relative filtering on the N-tuples of groups formed from
the N stream of groups passed on from the group-filter as input. The merger
rule module consists of a number of a submodules, where the output of the
merger is the set difference of the output of the first
submodule with the union of the output of the rest of the submodules. The
relative filtering on the groups are applied to express timing and concurrency
constraints using Allen interval algebra \cite{fallen:1983}

The ungrouper unwraps the tuples of group-records into individual
flow-records, ordered by their timestamps. The duplicate
flow-records appearing from several group-records are eliminated and are sent
as output only once.

>>>>

The flow query language pipeline consists of a number of independent elements
that are connected to one another using UNIX-based pipes.  Each element
receives the content from the previous pipe, performs an operation and pushes
it to the next element in the pipeline as shown in Fig.
\ref{fig:nfql-pipeline}. A complete description on the semantics of each
element in the pipeline can be found in \cite{vmarinov:2009}

The pipeline starts with the splitter. It takes the flow-records data as input
in the \texttt{flow-tools} compatible format. The splitter is responsible to
duplicate the input data out to several branches without any processing
whatsoever. This allows each of the branch to receive an identical copy of the
flow data to process it independently.

The filter is the first processing element of a branch in the pipeline.  It
performs \emph{absolute} filtering on the input flow-records data.  The
flow-records that pass the filtering criterion are forwarded to the grouper,
the rest of the flow-records are dropped. The filter compares separate fields
of a flow-record against either a constant value or a value on a different
field of the \emph{same} flow-record. The filter cannot \emph{relatively}
compare two different incoming flow-records

The grouper performs aggregation of the input flow-records data. It consists
of a number of terms that correspond to a specific subgroup. A flow-record in
order to be a part of the group should satisfy at-least one subgroup. A
flow-record can be a part of multiple subgroups within a group. A flow-record
cannot be part of multiple groups.  The grouping rules can be either absolute
or relative. The newly formed groups which are passed on to the group filter
can also contain meta-information about the flow-records contained within the
group using the aggregate clause defined as part of the grouper query.

The group-filter is the last processing element of the branch. It performs
\emph{absolute} filtering on the input group-records data. The group-records
that pass the filtering criterion are forwarded to the merger, the rest of the
group-records are dropped. The group-filter compares separate fields (or
aggregated fields) of a flow-record against either a constant value or a value
on a different field of the \emph{same} flow-record. The group-filter cannot
\emph{relatively} compare two different incoming group-records.

The merger performs relative filtering on the N-tuples of groups formed from
the N stream of groups passed on from the group-filter as input. The relative
filtering on the groups are applied to express timing and concurrency
constraints using Allen interval algebra \cite{fallen:1983}. The ungrouper is
the final processing element of the pipeline. It unwraps the tuples of
group-records into individual flow-records, ordered by their timestamps. The
duplicate flow-records appearing from several group-records are eliminated and
are sent as output only once.

### 2012-09-02 sections/evaluation.tex
* introduction   * capturing flows     * elevator pitch blurb     * netflow and ipfix     * wide applicability of flow capture techniques     * flow query language     * NFQL implementation     * paper organization * related work   * tcpdump, wireshark, flow-tools, nfdump, silk   * silk citation, simpleweb citation * flow query lanauage

We used a public flow trace containing $20M$ records to run a number of flow
queries using our benchmarking suite. We used trace $7$
\footnote{\url{http://traces.simpleweb.org/traces/netflow/netflow1/netflow000.tar.bz2}},
from Simpleweb \footnote{Simpleweb is a data repository of traffic traces from
University of Twente}.  The input trace was compressed at \texttt{ZLIB\_LEVEL}
$5$ using the \texttt{zlib} suite. It was also converted to \texttt{nfdump}
and SiLK compatible formats and compressed with \texttt{zlib}
keeping the same compression level.  The suite was run on a high-end
machine\footnote{\url{crystal.eecs.jacobs-university.de}} with $24$ cores of
$2.5$ GHz clock speed and $18$ GiB of memory.

>>>>

We used a public flow trace containing $20M$ records to run a number of flow
queries using our benchmarking suite. We used the publically available Trace 7
from SimpleWeb \cite{simpleweb} repository for all the performance
evaluations. The input trace was compressed at \texttt{ZLIB\_LEVEL} $5$ using
the \texttt{zlib} suite. It was also converted to \texttt{nfdump} and SiLK
compatible formats and compressed with \texttt{zlib} keeping the same
compression level. The suite was run on a high-end
machine\footnote{\url{crystal.eecs.jacobs-university.de}} with $24$ cores of
$2.5$ GHz clock speed and $18$ GiB of memory.

### 2012-09-02 sections/implementation/execution-workflow.tex
* introduction   * capturing flows     * elevator pitch blurb     * netflow and ipfix     * wide applicability of flow capture techniques     * flow query language     * NFQL implementation     * paper organization * related work   * tcpdump, wireshark, flow-tools, nfdump, silk   * silk citation, simpleweb citation * flow query lanauage

The flow query language suggests to read all the flow records of a supplied
trace into memory before starting the processing pipeline.  Since, the filter
stage uses the supplied set of absolute rules to make a decision on whether or
not to keep a flow record; it has to pass through the whole in-memory
recordset \emph{again} to fill in the filter results. This technique involves
multiple linear runs on the trace and therefore slows down when the ratio of
number of filtered records to the total number of flow-records is high. The
\ac{NFQL} implementation merges the filter stage with in-memory read of the
trace. This means, a decision on whether or not to make room for a record in
memory and eventually hold a pointer for it in filter results is done upfront
as soon as the record is read from the trace. In addition, if a request to
write the filter stage results to a flow-tools file has been made, the writes
are also made as soon the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. It is important to note that the filtered records are
saved in common location from where they are referenced by each branch. This
helps keep the memory costs at a minimum when multiple branches are involved.
A publically available Netflow $v5$ trace
\footnote{\url{http://traces.simpleweb.org/traces/netflow/netflow1/netflow000.tar.bz2}}
was used to compare the performance of the new filter in F$(v2)$ with that
from F$(v1)$ as shown in Fig. \ref{fig:fv1-fv2-filter}. The trace has around
$20M$ flow-records.

>>>>

The flow query language suggests to read all the flow records of a supplied
trace into memory before starting the processing pipeline.  Since, the filter
stage uses the supplied set of absolute rules to make a decision on whether or
not to keep a flow record; it has to pass through the whole in-memory
recordset \emph{again} to fill in the filter results. This technique involves
multiple linear runs on the trace and therefore slows down when the ratio of
number of filtered records to the total number of flow-records is high. The
\ac{NFQL} implementation merges the filter stage with in-memory read of the
trace. This means, a decision on whether or not to make room for a record in
memory and eventually hold a pointer for it in filter results is done upfront
as soon as the record is read from the trace. In addition, if a request to
write the filter stage results to a flow-tools file has been made, the writes
are also made as soon the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. It is important to note that the filtered records are
saved in common location from where they are referenced by each branch. This
helps keep the memory costs at a minimum when multiple branches are involved.
We used the publically available Trace 7 from the SimpleWeb \cite{simpleweb}
to compare the performance of the new filter in F$(v2)$ with that from F$(v1)$
as shown in Fig. \ref{fig:fv1-fv2-filter}. The trace has around $20M$
flow-records.

### 2012-09-02 sections/introduction.tex
* introduction   * capturing flows     * elevator pitch blurb     * netflow and ipfix     * wide applicability of flow capture techniques     * flow query language     * NFQL implementation     * paper organization * related work   * tcpdump, wireshark, flow-tools, nfdump, silk   * silk citation, simpleweb citation * flow query lanauage

Flow capture today, has emerged out to be one of the favored network
measurement techniques. This has largely been due to the reduction in the
monitoring traffic at the flow-level and the fine-grained control which was
not previously possible using SNMP interface-level queries. As a result,
each networking vendor has tried to come up with a standard protocol that
defines the semantics of this flow export. In this pursuit, Cisco eventually
managed to make their proprietary protocol so ubiquitously available, that the
next-generation universal standard is based on it.

\todo{...}

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\linewidth]{nfql-pipeline}
\caption{NFQL Pipeline \cite{vmarinov:2009}}
\label{fig:nfql-pipeline}
\end{figure*}

>>>>

Flow capture today, has emerged out to be one of the favored network
measurement techniques. In this technique, packets traversing a monitoring
point are not captured raw, instead they are aggregated together based on some
common characteristics. The common characteristics are learnt by inspecting
the packet headers as they cross the monitoring point. Flow-records resulting
from such an aggregation are then exported to a collector for further
analysis. This not only reduces the amount of traffic at the monitoring point,
but also provides fine-grained control over the network data which was not
previously possible using SNMP interface-level queries.

NetFlow and \ac{IPFIX} are the two popular standards of IP flow information
export. NetFlow \cite{rfc3954} is a proprietary network protocol designed by
Cisco Systems. It allow routers to generate and export flow records to a
designated collector. NetFlow v$9$ provides flexibility of user-tailored
export templates, \ac{MPLS} and IPv$6$ support and a larger set of flow keys.
\ac{IPFIX} \cite{rfc5101} on the other hand is an open standard by IETF deemed
to be the logical successor of NetFlow v$9$ on which it is based. The novelty
of the standard lies in its ability to describe record formats at runtime
using templates based on an extensible and well-defined information model. The
data transfer mechanism is also simplistic and extensible by being
unidirectional and transport protocol agnostic.

The wide applicability of this approach is easily seen from the pervasive use
of flow records for a vibrant set of network analysis applications. For
instance, the authors in \cite{mkim:2004} use the flow characteristics in the
traffic pattern to formalize a detection function that maps traffic patterns
to different DoS attacks, whereas in \cite{sdominik:2010}, the authors
use the flow-record data to exploit timing characteristics of webmail clients
to classify features that could identify webmail traffic from any other
traffic running over HTTPS.

It goes to show that understanding these intricate traffic patterns require
sophisticated flow analysis tools that can mine flow records for such a usage.
Unfortunately current tools fail to deliver owing to their poor language
design and simplistic filtering methods.  We recently proposed a flow query
language design \cite{vmarinov:thesis:2009} that aims to cater to such needs.
It can process flow records, aggregate them into groups, apply absolute (or
relative) filters and invoke Allen interval algebra rules \cite{fallen:1983}.

In this paper, we introduce \ac{NFQL}, an efficient C implementation of the
flow query language. It is the next iteration of the first prototype
implementation, Flowy \cite{kkanev:2010}, which was written in Python.
\ac{NFQL} however is not just a reimplementation in a new language, but
the inner workings have been reimagined to allow the execution engine to
scale to real-world sized traces. This has been possible by replacing the
performance hit stages of the pipeline with crispier algorithms that lower
down the execution times to factor making them comparable to contemporary
flow analysis tools.

The rest of the paper is organized as follows. In section \ref{sec:design} we
describe the flow query language by discussing each stage of the processing
pipeline. In section \ref{sec:implementation} we introduce the inner workings
of \ac{NFQL}. It begins by providing an overview of the \ac{NFQL} components
and the structure of the intermediate format used to exchange messages. The
workflow of the execution engine is described next and is supplemented by a
number of performance optimizations made to make the implementation scale. It
is followed by performance evaluation results by comparing them against
contemporary flow processing tools in section \ref{sec:evaluation} and
concludes with a discussion on its current limitation and future outlook in
section \ref{sec:conclusion}.

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\linewidth]{nfql-pipeline}
\caption{NFQL Processing Pipeline \cite{vmarinov:2009}}
\label{fig:nfql-pipeline}
\end{figure*}

### 2012-09-02 sections/relatedwork.tex
* introduction   * capturing flows     * elevator pitch blurb     * netflow and ipfix     * wide applicability of flow capture techniques     * flow query language     * NFQL implementation     * paper organization * related work   * tcpdump, wireshark, flow-tools, nfdump, silk   * silk citation, simpleweb citation * flow query lanauage

\texttt{tcpdump} and \texttt{wireshark} are the most popular tools used for
packet capture and analysis. \texttt{tcpdump} \cite{tcpdump-manpage} is a
premier command-line utility that uses the \texttt{libpcap}
\cite{pcap-manpage} library for packet capture. The power of \texttt{tcpdump}
comes from the richness of its expressions, the ability to combine them using
logical connectives and extract specific portions of a packet using filters.
\texttt{wireshark} \cite{wireshark-manpage} is a GUI application, aimed at
both journeymen and packet analysis experts. It supports a large number of
protocols, has a straightforward layout, excellent documentation, and can run
on all major operating systems.

\texttt{flow-tools} \cite{sromig:2000} is a suite of programs for capturing
and processing NetFlow v5 flow records. It consists of 24 separate tools that
work together by connecting them via UNIX pipes. \todo{...}

\texttt{nfdump} \cite{phaag:2006} works similar to flow-tools but uses a
different storage format. Flow records are captured using \texttt{nfcapd} and
then processed by \texttt{nfdump} which can filter as well as display the
sorted and filtered result. The power of its filtering rules is similar to
that of flow-tools and as such is mostly limited to absolute comparisons of
flow attributes.

SiLK\footnote{\url{http://tools.netsa.cert.org/silk/}} is a network traffic
collection and analysis tool developed and maintained by the CERT Network
Situational Awareness Team (CERT NetSA) at Carnegie Mellon University. SiLK is
the tool that comes quite close to providing similar capabilities as provided
by \ac{NFQL} and is therefore used as a reference point to compare the
performance of the \ac{NFQL} execution engine in this paper. The design and
implementation of SiLK, however, differs a lot from that of \ac{NFQL}.For
instance, in SiLK there are separate tools to perform the task of each stage
of the \ac{NFQL} processing pipeline. The stage functionality is not
full-fledged though. The grouping and merging operations can only be performed
using an equality operator. This is assumed in the tool, thereby allowing it
to perform optimization such as using hash tables to perform lookups. There
are also stringent requirements to how flow-data needs to be organized before
it can be piped into a tool. The grouping tool, for instance, assumes that the
to-be supplied input flow data is already sorted on the field column.  These
requirements make it a little cumbersome to design a full-fledged flow query.
For instance, trying to mimic a \ac{NFQL} query in SiLK ends up as a bash
script with over a dozen of SiLK tools piped together.

>>>>

The study of the network traffic patterns began with tools that utilized raw
packet captures at the monitoring point for traffic measurement.
\texttt{tcpdump} and \texttt{wireshark} have since then become the de-facto
tools for packet capture and analysis. \texttt{tcpdump} \cite{tcpdump-manpage}
is a premier command-line utility that uses the \texttt{libpcap}
\cite{pcap-manpage} library for packet capture. The power of \texttt{tcpdump}
comes from the richness of its expressions, the ability to combine them using
logical connectives and extract specific portions of a packet using filters.
\texttt{wireshark} \cite{wireshark-manpage} is a GUI application, aimed at
both journeymen and packet analysis experts. It supports a large number of
protocols, has a straightforward layout, excellent documentation, and can run
on all major operating systems.

In recent years, a number of tools have been developed that can capture the
traffic as flow records and use them for network analysis. \texttt{flow-tools}
and \texttt{nfdump} are one of the most popular tools used for analyzing
NetFlow data. \texttt{flow-tools} \cite{sromig:2000} is a suite of programs
for capturing and processing NetFlow v5 flow records. It consists of 24
separate tools that work together by connecting them via UNIX pipes. It can
capture, read, filter, and print flow records internally saved in a fixed-size
format. \texttt{nfdump} \cite{phaag:2006} is a very similar tool that uses a
different storage format. Flow records are captured using \texttt{nfcapd} and
then processed by \texttt{nfdump}. The power of filtering rules in both the
tools is however mostly limited to absolute comparisons of flow attributes. As
a result, relative comparison amongst different flows or querying a timing
relationship among them is not possible.  SiLK \cite{SiLK} is a network
traffic collection and analysis tool developed and maintained by the CERT
Network Situational Awareness Team (CERT NetSA) at Carnegie Mellon University.
SiLK is the tool that comes quite close to providing similar capabilities as
provided by \ac{NFQL} and is therefore used as a reference point to compare
the performance of the \ac{NFQL} execution engine in this paper. The design
and implementation of SiLK, however, differs a lot from that of \ac{NFQL}. For
instance, in SiLK there are separate tools to perform the task of each stage
of the \ac{NFQL} processing pipeline. The stage functionality is not
full-fledged though. The grouping and merging operations can only be performed
using an equality operator. This is assumed in the tool, thereby allowing it
to perform optimization such as using hash tables to perform lookups. There
are also stringent requirements to how flow-data needs to be organized before
it can be piped into a tool. The grouping tool, for instance, assumes that the
to-be supplied input flow data is already sorted on the field column.  These
requirements make it a little cumbersome to design a full-fledged flow query.
For instance, trying to mimic a \ac{NFQL} query in SiLK ends up as a bash
script with over a dozen of SiLK tools piped together.
